---
title: "Measures of Association"
author: "Mark Stevenson"
date: "`r Sys.Date()`"
bibliography: epiR_measures_of_association.bib
link-citations: yes
output:
  knitr:::html_vignette:
  toc: true
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{Measures of association}
  %\usepackage[utf8]{inputenc}

---
\setmainfont{Calibri Light}

```{r, echo = FALSE, message = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
options(tibble.print_min = 4L, tibble.print_max = 4L)
```

A common task in epidemiology is to quantify the strength of association between exposures ('risk factors') and disease outcomes. In this context the term 'exposure' is taken to mean a variable whose association with the outcome is to be estimated. 

Exposures can be harmful, beneficial or both harmful and beneficial (e.g. if an immunisable disease is circulating, exposure to immunising agents helps most recipients but may harm those who experience adverse reactions). The term 'outcome' is used to describe all the possible results that may arise from exposure to a causal factor or from preventive or therapeutic interventions [@porta_et_al:2008]. 

In this vignette we outline describe how `epiR` can be used to compute the various measures of association used in epidemiology notably the risk ratio, odds ratio, attributable risk in the exposed, attributable fraction in the exposed, attributable risk in the population and attributable fraction in the population. Examples are provided to demonstrate how the package can be used to deal with exposure-outcome data in various formats.  

Some preliminary comments. This vignette has been written assuming the reader routinely formats their 2 by 2 table data with the outcome status as columns and exposure status as rows. If this is not the case the argument `outcome = "as.columns"` (the default) can be changed to `outcome = "as.rows"`.

**Direct entry of 2 by 2 table contingency table cell frequencies**

Firstly, a 2 by 2 table can be created by listing the contingency table cell frequencies in vector format. Take the following example. 

A cross sectional study investigating the relationship between dry cat food (DCF) and feline lower urinary tract disease (FLUTD) was conducted [@willeberg:1977]. Counts of individuals in each group were as follows. DCF-exposed cats (cases, non-cases) 13, 2163. Non DCF-exposed cats (cases, non-cases) 5, 3349. Enter these data directly into R as a vector of length four. You can check that your counts are in the correct order by viewing the data as a matrix.

```{r}
dat.v01 <- c(13,2163,5,3349); dat.v01

# View the data in the usual 2 by 2 table format:
matrix(dat.v01, nrow = 2, byrow = TRUE)
```
Calculate the prevalence ratio, odds ratio, attributable prevalence in the exposed, attributable fraction in the exposed, attributable prevalence in the population and the attributable fraction in the population using function `epi.2by2`: 

```{r}
library(epiR)

epi.2by2(dat = dat.v01, method = "cross.sectional", conf.level = 0.95, units = 100, 
   interpret = FALSE, outcome = "as.columns")
```

The prevalence of FLUTD in DCF exposed cats was 4.01 (95% CI 1.43 to 11.23) times greater than the prevalence of FLUTD in non-DCF exposed cats.

In DCF exposed cats, 75% of FLUTD was attributable to DCF (95% CI 30% to 91%). Fifty-four percent of FLUTD cases in this cat population were  attributable to DCF (95% CI 4% to 78%).

Need a hand getting the correct wording to explain each of the listed measures of association and measures of effect? Set `interpret = TRUE` in `epi.2by2`:

```{r}
epi.2by2(dat = dat.v01, method = "cross.sectional", conf.level = 0.95, units = 100, 
   interpret = TRUE, outcome = "as.columns")
```

**Data frame with one row per observation**

Here we provide examples where you have exposure status and outcome status listed for each member of your study population. There are two options for contingency table preparation in this situation: (1) using base R's table function; or (2) using the `tidyverse` package.

For this example we use the low infant birth weight data presented by @hosmer_lemeshow:2000 and available in the `MASS` package in R. The `birthwt` data frame has 189 rows and 10 columns.  The data were collected at Baystate Medical Center, Springfield, Massachusetts USA during 1986. 

***Two by two table preparation using the table function in base R***

```{r}
library(MASS)

# Load and view the data:
dat.df02 <- birthwt; head(dat.df02)
```
Each row of this data set represents data for one mother. We're interested in the association between `smoke` (the mother's smoking status during pregnancy) and `low` (delivery of a baby less than 2.5 kg bodyweight). 

Its important that the table you present to `epi.2by2` is in the correct format: Outcome positives in the first column, outcome negatives in the second column, exposure positives in the first row and exposure negatives in the second row. If we run the `table` function on the `bwt` data the output table is in the wrong format:

```{r}
dat.tab02 <- table(dat.df02$smoke, dat.df02$low, dnn = c("Smoke", "Low BW")); dat.tab02
```
There are two ways to fix this problem. The quick fix is to simply ask R to switch the order of the rows and columns in the output table:

```{r}
dat.tab02 <- table(dat.df02$smoke, dat.df02$low, dnn = c("Smoke", "Low BW")); dat.tab02
dat.tab02 <- dat.tab02[2:1,2:1]; dat.tab02
```
The second approach is to set the exposure variable and the outcome variable as a factor and to define the levels of each factor using `levels = c(1,0)`:

```{r}
dat.df02$low <- factor(dat.df02$low, levels = c(1,0))
dat.df02$smoke <- factor(dat.df02$smoke, levels = c(1,0))
dat.df02$race <- factor(dat.df02$race, levels = c(1,2,3))

dat.tab02 <- table(dat.df02$smoke, dat.df02$low, dnn = c("Smoke", "Low BW")); dat.tab02
```
Now compute the odds ratio for smoking and delivery of a low birth weight baby:

```{r}
dat.epi02 <- epi.2by2(dat = dat.tab02, method = "cohort.count", conf.level = 0.95, 
   units = 100, interpret = FALSE, outcome = "as.columns")
dat.epi02
```
The odds of having a low birth weight child for smokers is 2.02 (95% CI 1.08 to 3.78) times greater than the odds of having a low birth weight child for non-smokers. 

***Two by two table preparation using tidyverse***

The `tidyverse` package can also be used to prepare data in the required format:

```{r}
library(tidyverse)

dat.df03 <- birthwt; head(dat.df03)

# Here we set the factor levels and tabulate the data in a single call using pipe operators:
dat.tab03 <- dat.df03 %>%
  mutate(low = factor(low, levels = c(1,0), labels = c("yes","no"))) %>%
  mutate(smoke = factor(smoke, levels = c(1,0), labels = c("yes","no"))) %>%
  group_by(smoke, low) %>%
  summarise(n = n()) 

# View the data:
dat.tab03

## View the data in conventional 2 by 2 table format:
pivot_wider(dat.tab03, id_cols = c(smoke), 
   names_from = low, values_from = n)
```
As before, compute the odds ratio for smoking and delivery of a low birth weight baby:

```{r}
dat.epi03 <- epi.2by2(dat = dat.tab03, method = "cohort.count", 
   conf.level = 0.95, units = 100, interpret = FALSE, outcome = "as.columns")
dat.epi03
```

The odds of having a low birth weight child for smokers is 2.02 (95% CI 1.08 to 3.78) times greater than the odds of having a low birth weight child for non-smokers. 

**Confounding**

We're concerned that the mother's race may confound the association between low birth weight and delivery of a low birth weight baby so we'll stratify the data by race and compute the Mantel-Haenszel adjusted odds ratio. As before, our tables can be prepared using either base R or `tidyverse`.

***Stratified two by two table preparation using the table function in base R***

```{r}
dat.df04 <- birthwt; head(dat.df04)

dat.tab04 <- table(dat.df04$smoke, dat.df04$low, dat.df04$race, 
   dnn = c("Smoke", "Low BW", "Race")); dat.tab04
```

Compute the Mantel-Haenszel adjusted odds ratio for smoking and delivery of a low birth weight baby, adjusting for the effect of race. Function `epi.2by2` automatically calculates the Mantel-Haenszel odds ratio and risk ratio when it is presented with stratified contingency tables.

```{r}
dat.epi04 <- epi.2by2(dat = dat.tab04, method = "cohort.count", 
   conf.level = 0.95, units = 100, interpret = FALSE, outcome = "as.columns")
dat.epi04
```

The Mantel-Haenszel test of homogeneity of the strata odds ratios is not significant (chi square test statistic 2.800; df 2; p-value = 0.25). We accept the null hypothesis and conclude that the odds ratios for each strata of race are the same. Because the stratum specific odds ratios are the same it is appropriate to compute a Mantel-Haenszel adjusted odds ratio.

After accounting for the confounding effect of race, the odds of having a low birth weight child for smokers was 3.09 (95% CI 1.49 to 6.39) times that of non-smokers.

***Stratified two by two table preparation using tidyverse***

```{r}
dat.df05 <- birthwt; head(dat.df05)

dat.tab05 <- dat.df05 %>%
  mutate(low = factor(low, levels = c(1,0), labels = c("yes","no"))) %>%
  mutate(smoke = factor(smoke, levels = c(1,0), labels = c("yes","no"))) %>%
  mutate(race = factor(race)) %>%
  group_by(race, smoke, low) %>%
  summarise(n = n()) 
dat.tab05

## View the data in conventional 2 by 2 table format:
pivot_wider(dat.tab05, id_cols = c(race, smoke), 
   names_from = low, values_from = n)
```

Compute the Mantel-Haenszel adjusted odds ratio for smoking and delivery of a low birth weight baby, adjusting for the effect of race:

```{r}
dat.epi05 <- epi.2by2(dat = dat.tab05, method = "cohort.count", 
   conf.level = 0.95, units = 100, interpret = FALSE, outcome = "as.columns")
dat.epi05
```

The Mantel-Haenszel test of homogeneity of the strata odds ratios is not significant (chi square test statistic 2.800; df 2; p-value = 0.25) so we accept the null hypothesis and conclude that the odds ratios for each strata of race are the same. After accounting for the confounding effect of race, the odds of having a low birth weight child for smokers is 3.09 (95% CI 1.49 to 6.39) times that of non-smokers.

Plot the individual strata odds ratios and the Mantel-Haenszel summary odds ratio as an error bar plot to better understand the process of adjustment.

```{r}
library(ggplot2); library(scales)

nstrata <- 1:length(unique(dat.tab05$race))
strata.lab <- paste("Strata ", nstrata, sep = "")
y.at <- c(nstrata, max(nstrata) + 1)
y.lab <- c("M-H", strata.lab)
x.at <- c(0.25,0.5,1,2,4,8,16,32)

or.p <- c(dat.epi05$massoc.detail$OR.mh$est, 
   dat.epi05$massoc.detail$OR.strata.cfield$est)
or.l <- c(dat.epi05$massoc.detail$OR.mh$lower, 
   dat.epi05$massoc.detail$OR.strata.cfield$lower)
or.u <- c(dat.epi05$massoc.detail$OR.mh$upper, 
   dat.epi05$massoc.detail$OR.strata.cfield$upper)
gdat.df05 <- data.frame(y.at, y.lab, or.p, or.l, or.u)

ggplot(data = gdat.df05, aes(x = or.p, y = y.at)) +
   geom_point() + 
   geom_errorbarh(aes(xmax = or.l, xmin = or.u, height = 0.2)) + 
   labs(x = "Odds ratio", y = "Strata") + 
   scale_x_continuous(trans = log2_trans(), breaks = x.at, 
      limits = c(0.25,32)) + 
   scale_y_continuous(breaks = y.at, labels = y.lab) + 
   geom_vline(xintercept = 1, lwd = 1) + 
   coord_fixed(ratio = 0.75 / 1) + 
   theme(axis.title.y = element_text(vjust = 0))
```

## References

---
nocite: '@*'
---